Depression and suicidality as credible signals of need

### 2021-05-22 Paper.rmd
* Tweaked emotion heatmaps * Added TOC * Removed two regression eqns * Minor edits * updated renv lockfile

---
title: "Depression and suicidality as credible signals of need"
author: "Michael R. Gaffney, Kai Adams, Kristen L. Syme, and Edward H. Hagen"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
always_allow_html: false
output: 
  bookdown::html_document2:
    number_sections: false
bibliography: MTBib.bib
csl: apa-6th-edition.csl
---

@@

Almost all individuals will suffer adversity, such as injury or loss of material or social resources, at some point in their lives. Psychological pain, such as sadness and low mood, probably evolved to motivate victims of adversity to shift their attention to the causes of adversity so as to mitigate its negative fitness consequences and to learn to avoid future such adverse events [@thornhill1989; @andrews2010; @Nesse1990]. In many cases, social partners could help, and therefore signals of psychological pain, such as sad expressions and crying, probably evolved to indicate need [e.g., @balsters2013].

@@

There is strong evidence that most cases of depression are caused by major negative life events [@Hammen2005; @Mazure1998; @devries2011; @ellsberg2008]. Compared to non-depressed individuals, those with depression report about twice as many instances of negative events than non-depressed individuals [@Mazure1998] and more negative events than those with schizophrenia and bipolar depression across multiple studies [@Paykel1994]. In longitudinal studies, depression onsets soon after a negative event [@KendlerEtAl1999] or coincides with periods where adversity is likely to increase prior to it [e.g., depression starting before a divorce rather than after; @rosenstrom2017]. Additional support for the causal role of negative life events comes from the fact that the relationship between negative events and depression holds true when looking at events outside of one's control, showing that the connection is unlikely to be driven solely by individuals who are already depressed creating situations where negative events are likely to be common [@Hammen2005; @KendlerEtAl1999]. Furthermore, twin studies have shown that one's history of negative events is still a strong predictor of major depression when controlling for genetic similarity, and that part of the heritability of depression stems from the heritability of negative events like divorce and family conflict [@KendlerBaker2007; @KendlerEtAl1999]. For these and other reasons, most cases of depression are probably not due to mental illness but are instead functional instances of psychological pain [@Hagen:1999; @Hagen2003; @horwitz_loss_2007; @andrews2010; @SymeHagen2021].

@@

The primary aim of this study was to examine responses to emotional signals that varied from low to high cost, in an experimental vignette study in which a possible victim of adversity asks for help from the participant, but has incentives to exaggerate her need. In the bargaining framework, the willingness to delay credibly signals the valuation of current cooperative arrangements, with greater delays indicating lower valuations. As signal cost increased, we predicted that participants would report (1) increased belief in the signaler's claims and (2) increased likelihood of providing help, with (3) the increased likelihood of providing help mediated by the increased belief in the signaler's need.

@@

After rating their beliefs and actions, and which emotions they thought the potential victim was experiencing, participants were randomized into either the control condition or one of four emotional signals by the victim (in order of increasing signal cost): (1) control condition: a verbal request without additional signaling; (2) crying; (3) mild-depression; (4) depression; and (5) a suicide attempt. The signals involved the participant encountering the victim some time after the adverse event and observing, e.g., crying, sad expressions, or poor personal hygiene. These descriptions did not use the terms depression, depressed, suicidal, or mental health.

The signals were cumulative: crying can be an important feature of depression [for discussion on the relationship between crying and depression see @bylsma2020], and depression is a major risk factor for suicide [@BostwickPankratz2000; @Kessler2012]. Accordingly, components of less-costly signals were included in more-costly signals. Although we use the term 'signals' throughout the paper for brevity, we expect that the hypothesized signals of need, like many signals, would also provide information to others in the form of cues [for discussion on the evolution of signals from cues see: @biernaskie2018; @Tiokhin2016; @Steinkopf2015]. The full text for the signals are available in the SI.

@@

The proportion of participants inferring each emotion in each vignette at T1 is depicted in Figure \@ref(fig:t1emotionheatmap), after each signal at T2 in Figure \@ref(fig:t2emotionheatmap), and the change in proportions from T1 to T2 in Figure \@ref(fig:diffemotions). The increase in inferred depression in the *Depression* and *Suicide attempt* conditions, but not the *Verbal request* and *Crying* conditions, helps validate these signals, as does the increase in inferred suicidality in the *Suicide attempt* condition.

@@

The `glm` models we report in the main text were:

`r extract_eq(models$Model$m1, intercept='beta', wrap=T)` `r extract_eq(models$Model$m3, intercept='beta', wrap=T)` `r extract_eq(models$Model$m1b, intercept='beta', wrap=T)` `r extract_eq(models$Model$m3b, intercept='beta', wrap=T)`

@@

## T1 emotion vs vignette heatmap

```{r t1emotionheatmap, fig.width=8, fig.cap="Time 1 inference of victim's emotional state in each vignette. Each cell represents the proportion of participants who inferred that emotion. Participants could check multiple emotions from the list. Rows and columns ordered with the PCA angle method in the seriation package."}
hagenheat(dfT1emotion_vignette, rotate_labels = F, seriation_method = 'PCA_angle', viridis_option = 'B') + theme_minimal(15)
```

## T2 emotion vs signal heatmap

@@

hagenheat(dfT2emotion_signal, rotate_labels = F, seriation_method = 'PCA_angle', viridis_option = 'B') + theme_minimal(15)
```

@@

hagenheat(d_diff, rotate_labels = F, seriation_method = 'PCA_angle') + scale_fill_gradient2(low='blue', high='red')

@@

@@

>>>>

---
title: "Depression and suicidality as credible signals of need"
author: "Michael R. Gaffney, Kai Adams, Kristen L. Syme, and Edward H. Hagen"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
always_allow_html: false
output: 
  bookdown::html_document2:
    toc: true
    number_sections: false
bibliography: MTBib.bib
csl: apa-6th-edition.csl
---

@@

Almost all individuals will suffer adversity, such as injury or loss of material or social resources, at some point in their lives. Psychological pain, such as sadness and low mood, probably evolved to motivate victims of adversity to shift their attention to the causes of adversity so as to mitigate its negative fitness consequences and to learn to avoid future such adverse events [@thornhill1989; @andrews2010; @Nesse1990]. In many cases, social partners could help victims, and therefore signals of psychological pain, such as sad expressions and crying, probably evolved to indicate need [e.g., @balsters2013].

@@

There is strong evidence that most cases of depression are caused by major negative life events [@Hammen2005; @Mazure1998; @devries2011; @ellsberg2008]. Compared to non-depressed individuals, those with depression report about twice as many instances of negative events than non-depressed individuals [@Mazure1998] and more negative events than those with schizophrenia and bipolar depression across multiple studies [@Paykel1994]. In longitudinal studies, depression onsets soon after a negative event [@KendlerEtAl1999] or coincides with periods where adversity is likely to increase prior to it [e.g., depression starting before a divorce rather than after, @rosenstrom2017]. Additional support for the causal role of negative life events comes from the fact that the relationship between negative events and depression holds true when looking at events outside of one's control, showing that the connection is unlikely to be driven solely by individuals who are already depressed creating situations where negative events are likely to be common [@Hammen2005; @KendlerEtAl1999]. Furthermore, twin studies have shown that one's history of negative events is still a strong predictor of major depression when controlling for genetic similarity, and that part of the heritability of depression stems from the heritability of negative events like divorce and family conflict [@KendlerBaker2007; @KendlerEtAl1999]. For these and other reasons, most cases of depression are probably not due to mental illness but are instead functional instances of psychological pain [@Hagen:1999; @Hagen2003; @horwitz_loss_2007; @andrews2010; @SymeHagen2021].

@@

In the bargaining framework, the willingness to delay credibly signals the valuation of current cooperative arrangements, with greater delays indicating lower valuations. The primary aim of this study was to examine responses to emotional signals that varied in the extent to which they reduced productivity, which we refer to as costs, in an experimental vignette study in which a possible victim of adversity asks for help from the participant, but has incentives to exaggerate her need. As signal cost increased, we predicted that participants would report (1) increased belief in the signaler's claims and (2) increased likelihood of providing help, with (3) the increased likelihood of providing help mediated by the increased belief in the signaler's need.

@@

After rating their beliefs and actions, and which emotions they thought the potential victim was experiencing, participants were randomized into either the control condition or one of four emotional signals by the victim (in order of increasing signal cost): (1) control condition: a verbal request without additional signaling; (2) crying; (3) mild-depression; (4) depression; and (5) a suicide attempt. The signals involved the participant encountering the victim some time after the adverse event and observing, e.g., crying; sad expressions; reduced effort, fatigue, and poor personal hygiene; and suicidal self-injury. These descriptions did not use the terms depression, depressed, suicidal, or mental health.

The signals were cumulative: crying can be an important feature of depression [for discussion on the relationship between crying and depression see @bylsma2020], and depression is a major risk factor for suicide [@BostwickPankratz2000; @Kessler2012]. Accordingly, components of less-costly signals were included in more-costly signals. Although we use the term 'signals' throughout the paper for brevity, we expect that the hypothesized signals of need, like many signals, would also provide information to others in the form of cues [for discussion on the evolution of signals from cues see: @biernaskie2018; @Tiokhin2016; @Steinkopf2015]. The complete texts of the signals are available in the SI.

@@

The proportion of participants inferring each emotion in each vignette at T1 and after each signal at T2 is depicted in Figure \@ref(fig:emotionheatmaps), and the change in proportions from T1 to T2 in Figure \@ref(fig:diffemotions). The increase in inferred depression in the *Depression* and *Suicide attempt* conditions, but not the *Verbal request* and *Crying* conditions, helps validate these signals, as does the increase in inferred suicidality in the *Suicide attempt* condition.

@@

The `glm` models we report in the main text:

`r extract_eq(models$Model$m1, intercept='beta', wrap=T)` 
`r extract_eq(models$Model$m3, intercept='beta', wrap=T)`

@@

## Emotions in vignette (T1) and signal (T2) conditions

```{r emotionheatmaps, fig.width=10, fig.cap="Each cell represents the proportion of participants who inferred that emotion. **Time 1 emotions**: inference of victim's emotional state in each vignette at time 1. **Time 2 emotions**: inference of victim's emotional state following the signal at time 2. Participants could check multiple emotions from the list. Rows ordered by mean proportion across vignettes."}

row_order = rowSums(dfT1emotion_vignette[-1])

dfT1emotion_vignette$name <- factor(dfT1emotion_vignette$name, levels = dfT1emotion_vignette$name[order(row_order)])

values <- seq(0, 1, 0.25)

p1 <- 
  dfT1emotion_vignette %>% 
  gather(key = key, value = value, -1) %>%
  mutate(key = factor(key, levels = c('Thwarted marriage', 'Basketball coach', 'Romantic partner', 'Brother-in-law'))) %>%
  ggplot(aes_string("key", "name", fill = "value")) +
  geom_tile() + 
  scale_fill_viridis_c('Proportion', option = 'B', limits = c(0, 1), breaks = values) +
  scale_x_discrete(labels = scales::label_wrap(10)) +
  labs(title = 'T1 emotions', x = "", y = "") + 
  theme_minimal()

dfT2emotion_signal$name <- factor(dfT2emotion_signal$name, levels = dfT1emotion_vignette$name[order(row_order)])

p2 <- 
  dfT2emotion_signal %>% 
  gather(key = key, value = value, -1) %>%
  mutate(key = factor(key, levels = c('Verbal request', 'Crying', 'Mild depression', 'Depression', 'Suicide attempt'))) %>%
  ggplot(aes_string("key", "name", fill = "value")) +
  geom_tile() + 
  scale_fill_viridis_c('Proportion', option = 'B', limits = c(0, 1), breaks = values) +
  scale_x_discrete(labels = scales::label_wrap(10)) +
  labs(title = 'T2 emotions', x = "", y = "") + 
  theme_minimal() +
  theme(axis.text.y = element_blank())

p1 + p2 + plot_layout(guides='collect')
```

```{r t1emotionheatmap, fig.width=8, fig.cap="Time 1 inference of victim's emotional state in each vignette. Each cell represents the proportion of participants who inferred that emotion. Participants could check multiple emotions from the list. Rows and columns ordered with the PCA angle method in the seriation package."}
# hagenheat(dfT1emotion_vignette, rotate_labels = F, seriation_method = 'PCA_angle', viridis_option = 'B') + theme_minimal(15)
```

<!-- ## T2 emotion vs signal -->

@@

# hagenheat(dfT2emotion_signal, rotate_labels = F, seriation_method = 'PCA_angle', viridis_option = 'B') + theme_minimal(15)
```

@@

hagenheat(d_diff, rotate_labels = F, seriation_method = 'PCA_angle', rev_col = T) + scale_fill_gradient2(low='blue', high='red')

@@

At the end of the study, participants could optionally provide written feedback about study. We classified all feedback into two categories: *Generic* for a few short words, such as 'nice study', or 'thanks'; and *Informative* for one or two sentences commenting on specific aspects of the survey. See Figure \@ref(fig:feedbackplot).

@@

We then further classified the *Informative* feedback into six categories: the vignette corresponds to real-life experiences; comments on the vignette (e.g., other ways to respond); what the participant thought about the T3 information that the victim was telling the truth; criticisms of the study (e.g., the attention checks); comments on the study; and an explanation of why the participant acted or failed to act. See Figure \@ref(fig:informativeplot). The full texts of all comments are in Table \@ref(tab:informativetab).
